The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-17 rnw/Analysis and Visualization.Rnw
update Andreas 11.08.2017(2)
update Andreas 17.08.2017

\section{Special Cases and Scenarios}
\label{sec:speccas_and_scen}

>>>>

\section{Analysis and Visualization}
\label{sec:anal_and_vis}

% include:
% - estTable()-function demo
% - mphase.gain()-function demo
% - plot.esttable()-function demo
% - 3-4 pages max
% - pronounce that "it is crucial to quantify the potential benefit of the mphase procedures
%   compared to the onephase procedure --> allows to calculate the financial benefit --> rel.eff

The package \pkg{forestinventory} also provides methods for \textit{analysing} and \textit{visualizing} the estimation results created by the \code{onephase}, \code{twophase} and \code{threephase} functions.

\subsection{Analysis}

If we want to compare the results and performance of different estimation methods and estimators for a given global or small area inventory data set, the function to start with is \code{estTable()}. This function restructures the results of \code{onephase}, \code{twophase} and \code{threephase} estimations and merges them into one single data set that provides the basis for further analysis. For demonstration purposes, we will first recalculate the one-phase estimations as well as the two-phase and three-phase \textit{extended pseudo synthetic} and the \textit{pseudo synthetic} estimations for the \code{grisons} dataset:

\begin{small}
<<>>=
op <- onephase(formula = tvol~1, data = grisons,
               phase_id = list(phase.col = "phase_id_2p", terrgrid.id = 2),
               area = list(sa.col = "smallarea", areas = c("A", "B", "C", "D")))
@
\end{small}

\begin{small}
<<>>=
extpsynth_2p<- twophase(formula = tvol ~ mean + stddev + max + q75, data = grisons,
                        phase_id = list(phase.col = "phase_id_2p", terrgrid.id = 2),
                        small_area = list(sa.col = "smallarea", 
                                          areas = c("A", "B","C", "D"),
                                          unbiased = TRUE))
@
\end{small}

\begin{small}
<<>>=
psynth_2p <- twophase(formula = tvol ~ mean + stddev + max + q75, data = grisons,
                      phase_id = list(phase.col = "phase_id_2p", terrgrid.id = 2),
                      boundary_weights = "boundary_weights",
                      small_area = list(sa.col = "smallarea", 
                                        areas = c("A", "B", "C", "D"),
                                        unbiased = FALSE))
@
\end{small}

\begin{small}
<<>>=
extpsynth_3p<- threephase(formula.s0 = tvol ~ mean,
                formula.s1 = tvol ~ mean + stddev + max + q75, data = grisons,
                phase_id = list(phase.col = "phase_id_3p", s1.id = 1, terrgrid.id = 2),
                small_area=list(sa.col = "smallarea", areas = c("A", "B", "C", "D"),
                                unbiased = TRUE))
@
\end{small}

\begin{small}
<<>>=
psynth_3p<- threephase(formula.s0 = tvol ~ mean,
                formula.s1 = tvol ~ mean + stddev + max + q75, data = grisons,
                phase_id = list(phase.col = "phase_id_3p", s1.id = 1, terrgrid.id = 2),
                small_area = list(sa.col = "smallarea", 
                                  areas = c("A", "B", "C", "D"),
                                  unbiased = FALSE))
@
\end{small}

We can then aggregate all estimation objects in a \code{list} and pass it to the \code{estTable}-function:

\begin{small}
<<>>=
grisons.sae.table<- estTable(est.list = list(op, extpsynth_2p, psynth_2p,
                                             extpsynth_3p, psynth_3p),
                             sae = TRUE, 
                             vartypes = c("variance", "g_variance",  "ext_variance"))
@
\end{small}

The function merges the estimation results and returns a \code{list} object with the subclasses \code{"esttable"  "smallarea"}. If one prefers the \code{data.frame} format for further analysis, this can easily be done using \code{as.data.frame(grisons.sae.table)}. Note however that the package \pkg{forestinventory} provides several \code{S3} generic functions that particularly respond to the class \code{"esttable"}. Note that the \code{vartypes} argument can be used to restrict the \code{estTable} output to certain types of variances.

The structure of an \code{esttable} object is very similar to the objects created by the small area estimation functions of the package. However, the point estimates and variances from all estimation objects passed to \code{estTable} have been stored in one single column (\code{estimate} and \code{variance}) and can be distinguished by the variables \code{method}, \code{estimator} and \code{vartype} which specify the estimation method (one, two or threephase), the estimator and the type of variance that was applied.

\begin{small}
<<eval=FALSE>>=
str(grisons.sae.table)
@
\end{small}

\begin{footnotesize}
<<echo=FALSE>>=
str(grisons.sae.table)
@
\end{footnotesize}

Note that \code{estTable} also returns the \textbf{\textit{estimation error}} (\code{error}) that is defined as the \textit{standard error} devided by the \textit{point estimate}:

\begin{equation}\label{eq:esterror}
  error_{[\%]}=\frac{\sqrt{\hat{\var}(\hat{Y})}}{\hat{Y}}\cdot100
\end{equation}

As mentioned in the introduction (section \ref{sec:intro}), the application of multi-phase estimation techniques in forest practice have two main \textbf{goals}:  
\begin{enumerate}
  \item \textit{\textbf{maintaining}} the one-phase estimation precision under reduced costs, i.e. reduced $s_2$ sample size
  \item \textit{\textbf{increasing}} the estimation precision at identical costs, i.e. without increasing the $s_2$ sample size
\end{enumerate}

The function \code{mphase.gain()} was particularly written to quantify the potential benefit of a multi-phase global or small area estimation compared to its respective one-phase estimation. The function takes an \code{esttable} object as input and returns a validation of which multi-phase method and estimator performed best in comparison to the onephase estimation (\textit{baseline}) in terms of estimation precision. If the \code{esttable} object contains more than one multi-phase estimation object, \code{mphase.gain} identifies the one with the smallest variance and compares it to the \code{onephase} estimation. The argument \code{pref.vartype} can be used to define what type of variance (\textit{g-weight} or \textit{external}) should be used for the comparison. Per default (\code{exclude.synth = TRUE}), synthetic estimations (\textit{synth} and \textit{psynth} estimator) are not considered for the comparison since they normally have a much smaller variance but to the cost of a potential bias.

<<echo=FALSE>>=
options(width=10000) 
@

\begin{small}
<<>>=
mphase.gain(grisons.sae.table)
@
\end{small}

% quantify the potential benefit of the mphase compared to the onephase estimation method
% - pronounce that "it is crucial to quantify the potential benefit of the mphase procedures
%   compared to the onephase procedure --> allows to quantify the financial benefit --> rel.eff

### 2017-08-17 rnw/Twophase_Estimators and their Application.Rnw
update Andreas 17.08.2017

This procedure is identical under both the \textit{exhaustive} and \textit{non-exhaustive} derivation of the explanatory variables. Based on the regression coefficients, the prediction for the sample location $x$ is calcualted as $\pmb{Z}^t(x)\hat{\pmb{\beta}}_{s_2}$ and the empirical model residuals, which are only available at all sample locations $x \in s_2$, are calculated as $\hat{R}(x)=Y(x)-\hat{Y}(x)$. Note that in the \proglang{R} package \pkg{forestinventory}, only \textit{internal models} can be used to calculate estimations. This means that the regression model and thus the vector of regression coefficients is always derived from the current inventory data that are passed to the \code{twophase} and \code{threephase} function as input data. \citet{massey2015a} also describe the use of external models in the design based framework where the regression coefficients can be taken e.g. from previous studies. However, a respective function in the package is up to now non existing and will be subject to future extentions.\par

@@

The package \pkg{forestinventory} offers 3 small area estimators. A general principle of all small area estimators of the package is to restrict the information to the small area $G$ whenever possible. An exception however is the retrieval of the regression coefficient vector $\hat{\pmb{\beta}}$ and its variance-covariance matrix $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$, which is always calculated using the entire sample $s_2$ according to equation \ref{normequ_simple} and \ref{eq:estvarmatrix}.\par

@@

We first introduce the \textbf{small area estimator} (\textit{small}), which uses exhaustively computed explanatory variables, and its non-exhaustive version, the \textbf{\textit{pseudo}} \textbf{small area estimator} (\textit{psmall}). Their point estimates are given in equation \ref{eq:pointest_2p_small} and \ref{eq:pointest_2p_psmall}. As obvious, the globally derived regression coefficients are first applied to the exhaustively or non-exhaustively calculated means of the explanatory variables ($\bar{\pmb{Z}}_G$, $\hat{\bar{\pmb{Z}}}_G$) available at the sample location $s_{1,G}$ in the small area. A potential bias of the regression model predictions in the small area $G$ is then corrected by adding the mean of the residuals in $G$ to the first term. We call this the \textit{bias} or \textit{residual correction} term.

\begin{subequations}\label{eq:pest_2p_small_psmall}
\begin{align}
  \hat{Y}_{G,small,2p} & =\bar{\pmb{Z}}_G^t\hat{\pmb{\beta}}_{s_2} + \frac{1}{n_{2,G}}\hat{R}(x)  \label{eq:pointest_2p_small} \\
  \hat{Y}_{G,psmall,2p} & =\hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\beta}}_{s_2} + \frac{1}{n_{2,G}}\hat{R}(x) \label{eq:pointest_2p_psmall}
\end{align}
\end{subequations}

The package again provides the \textit{design-based} (equation \ref{eq:var_2p_reg_small}, \ref{eq:var_2p_reg_psmall}) and \textit{external} variance (equation \ref{eq:varext_2p_reg_small}, \ref{eq:varext_2p_reg_psmall}) for the \textit{small} and the \textit{psmall} estimator. Again note that all components are restricted to those available at the sample locations $s_{1,G}$ and $s_{2,G}$ in the small area, with exception of the regression coeffient components $\hat{\pmb{\beta}}_{s_2}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$.

@@

The \textbf{synthetic} and \textbf{pseudo synthetic estimator} is applied if no terrestrial samples are available within the small area $G$ (i.e. $n_{2,G}=0$). In this case, the point estimate (equation \ref{eq:pointest_2p_reg_synth} and \ref{eq:pointest_2p_reg_psynth}) is based \textit{only} on the predictions generated by applying the globally derived regression model to the auxiliary vector $\bar{\pmb{Z}}_G$ or $\hat{\bar{\pmb{Z}}}_G$. However, the bias correction applied in the small and pseudo small area estimator (equation \ref{eq:pointest_2p_small} and \ref{eq:pointest_2p_psmall}) is no longer possible. Thus, the synthetic / pseudo synthetic estimator is unbiased \textbf{only} if the \textit{zero mean property} of the residuals $\int_F R(x)dx=0$ also holds for the residuals in the small area $G$, i.e. $\int_G R(x)dx=0$. If this assumption is not fullfilled, the synthetic point estimate $\hat{Y}_{G,synth,2p}$ and the pseudo synthetic point estimate $\hat{Y}_{G,psynth,2p}$ will have a design-based asymptotic bias equal to $-\frac{1}{\lambda(G)}\int_G R(x)$. Also note that residual variation can no longer be considered in the design-based variance (equation \ref{eq:var_2p_reg_synth} and \ref{eq:var_2p_reg_psynth}). Therefore, the synthetic estimators will usually have a smaller variance than estimators incorporating the regression model uncertainties, but at the cost of a potential bias. Due to the absence of available residuals in $G$, there is also no external variance form for the synthetic and pseudo synthetic estimator.

@@

In the \textit{small} and \textit{psmall} estimator, the crucial zero mean residual property for the small area $G$ is forced by the bias adjustment terms (equation \ref{eq:pointest_2p_small}, \ref{eq:pointest_2p_psmall}).  However, the equations would considerably simplify if we could mathematically insure that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This idea is realized in the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) by extending the vector of auxiliary information $\pmb{Z}(x)$ by a binary categorical indicator variable $I_G(x)$ which takes the value \code{1} if the observation lies inside the small area $G$ and is otherwise set to \code{0}. The \textit{extended} auxiliary vector thus reads as $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and leads to a classical ANOVA regression model. Note that the intercept of the regression model is thus adjusted for all data points belonging to small area $G$ which insures that the mean of the residuals $\frac{1}{n_{2,G}}\sum_{x\in{s_{2,G}}}R(x)$ in $G$ are always zero by construction.\par

@@

\begin{equation}\label{ext_estvarcovaux_G}
\hat{\Sigma}_{\hat{\bar{\pmb{\mathbb{Z}}}}_{1,G}}=
\frac{1}{n_{1,G} (n_{1,G}-1)}\sum_{x\in{s_{1,G}}}
(\pmb{\mathbb{Z}}(x)-\hat{\bar{\pmb{\mathbb{Z}}}}_{1,G})(\pmb{\mathbb{Z}}(x)-
\hat{\bar{\pmb{\mathbb{Z}}}}_{1,G})^t
\end{equation}

@@

The estimators are called \textit{synthetic} since no residual term is present such as in the \textit{small} and \textit{psmall} estimator (equation \ref{eq:pointest_2p_small} and \label{eq:pointest_2p_psmall}) and thus look similar to the synthetic estimators (equation \ref{eq:pointest_2p_reg_synth}, \ref{eq:pointest_2p_reg_synth}, \ref{eq:var_2p_reg_synth}, \ref{eq:var_2p_reg_psynth}). However, note that by a decomposition of $\hat{\pmb{\theta}}_{s_2}$ it can be shown  that the residual correction is indeed included in the regression coefficient $\hat{\pmb{\theta}}_{s_2}$ \citep{mandallaz2016} and thus the estimator is design-unbiased.

>>>>

This procedure is identical under both the \textit{exhaustive} and \textit{non-exhaustive} derivation of the explanatory variables. Based on the regression coefficients, the prediction for the sample location $x$ is calculated as $\pmb{Z}^t(x)\hat{\pmb{\beta}}_{s_2}$ and the empirical model residuals, which are only available at all sample locations $x \in s_2$, are calculated as $\hat{R}(x)=Y(x)-\hat{Y}(x)$. Note that in the \proglang{R} package \pkg{forestinventory}, only \textit{internal models} can be used to calculate estimations. This means that the regression model and thus the vector of regression coefficients is always derived from the current inventory data that are passed to the \code{twophase} and \code{threephase} function as input data. \citet{massey2015a} also describe the use of external models in the design based framework where the regression coefficients can be taken e.g. from previous studies. However, a respective function in the package is up to now non existing and will be subject to future extentions.\par

@@

The package \pkg{forestinventory} offers 3 small area estimators. A general principle of all small area estimators of the package is to restrict the information to the small area $G$ whenever possible. An exception however is the retrieval of the regression coefficient vector $\hat{\pmb{\beta}}$ and its variance-covariance matrix $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$, which are always calculated using the entire sample $s_2$ according to equation \ref{normequ_simple} and \ref{eq:estvarmatrix}.\par

@@

We first introduce the \textbf{small area estimator} (\textit{small}), which uses exhaustively computed explanatory variables, and its non-exhaustive version, the \textbf{\textit{pseudo}} \textbf{small area estimator} (\textit{psmall}). Their point estimates are given in equation \ref{eq:pointest_2p_small} and \ref{eq:pest_2p_psmall}. As obvious, the globally derived regression coefficients are first applied to the exhaustively or non-exhaustively calculated means of the explanatory variables ($\bar{\pmb{Z}}_G$, $\hat{\bar{\pmb{Z}}}_G$) available at the sample location $s_{1,G}$ in the small area. A potential bias of the regression model predictions in the small area $G$ is then corrected by adding the mean of the residuals in $G$ to the first term. We call this the \textit{bias} or \textit{residual correction} term.

\begin{subequations}\label{eq:pest_2p_small_psmall}
\begin{align}
  \hat{Y}_{G,small,2p} & =\bar{\pmb{Z}}_G^t\hat{\pmb{\beta}}_{s_2} + \frac{1}{n_{2,G}}\hat{R}(x)  \label{eq:pointest_2p_small} \\
  \hat{Y}_{G,psmall,2p} & =\hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\beta}}_{s_2} + \frac{1}{n_{2,G}}\hat{R}(x) \label{eq:pest_2p_psmall}
\end{align}
\end{subequations}

The package again provides the \textit{design-based} (equation \ref{eq:var_2p_reg_small}, \ref{eq:var_2p_reg_psmall}) and \textit{external} variance (equation \ref{eq:varext_2p_reg_small}, \ref{eq:varext_2p_reg_psmall}) for the \textit{small} and the \textit{psmall} estimator. Again note that all components are restricted to those available at the sample locations $s_{1,G}$ and $s_{2,G}$ in the small area, with exception of the regression coefficient components $\hat{\pmb{\beta}}_{s_2}$ and $\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}$.

@@

The \textbf{synthetic} and \textbf{pseudo synthetic estimator} is applied if no terrestrial samples are available within the small area $G$ (i.e. $n_{2,G}=0$). In this case, the point estimate (equation \ref{eq:pointest_2p_reg_synth} and \ref{eq:pointest_2p_reg_psynth}) is based \textit{only} on the predictions generated by applying the globally derived regression model to the auxiliary vector $\bar{\pmb{Z}}_G$ or $\hat{\bar{\pmb{Z}}}_G$. However, the bias correction applied in the small and pseudo small area estimator (equation \ref{eq:pointest_2p_small} and \ref{eq:pest_2p_psmall}) is no longer possible. Thus, the (pseudo) synthetic estimator is unbiased \textbf{only} if the \textit{zero mean property} of the residuals $\int_F R(x)dx=0$ also holds for the residuals in the small area $G$, i.e. $\int_G R(x)dx=0$. If this assumption is not fullfilled, the synthetic point estimate $\hat{Y}_{G,synth,2p}$ and the pseudo synthetic point estimate $\hat{Y}_{G,psynth,2p}$ will have a design-based asymptotic bias equal to $-\frac{1}{\lambda(G)}\int_G R(x)$. Also note that the residual variation can no longer be considered in the design-based variance (equation \ref{eq:var_2p_reg_synth} and \ref{eq:var_2p_reg_psynth}). Therefore, the synthetic estimators will usually have a smaller variance than estimators incorporating the regression model uncertainties, but at the cost of a potential bias. Due to the absence of available residuals in $G$, there is also no external variance form for the synthetic and pseudo synthetic estimator.

@@

In the \textit{small} and \textit{psmall} estimator, the crucial zero mean residual property for the small area $G$ is forced by the bias adjustment terms (equation \ref{eq:pointest_2p_small}, \ref{eq:pest_2p_psmall}).  However, the equations would considerably simplify if we could mathematically insure that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This idea is realized in the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) by extending the vector of auxiliary information $\pmb{Z}(x)$ by a binary categorical indicator variable $I_G(x)$ which takes the value \code{1} if the observation lies inside the small area $G$ and is otherwise set to \code{0}. The \textit{extended} auxiliary vector thus reads as $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and leads to a classical ANOVA regression model. Note that the intercept of the regression model is thus adjusted for all data points belonging to small area $G$ which insures that the mean of the residuals $\frac{1}{n_{2,G}}\sum_{x\in{s_{2,G}}}R(x)$ in $G$ are always zero by construction.\par

@@

\begin{equation}\label{ext_estvarcovaux_G}
\hat{\Sigma}_{\hat{\bar{\pmb{\mathbb{Z}}}}_{1,G}}=
\frac{1}{n_{1,G} (n_{1,G}-1)}\sum_{x\in{s_{1,G}}}
(\pmb{\mathbb{Z}}(x)-\hat{\bar{\pmb{\mathbb{Z}}}}_{G})(\pmb{\mathbb{Z}}(x)-
\hat{\bar{\pmb{\mathbb{Z}}}}_{G})^t
\end{equation}

@@

The estimators are called \textit{synthetic} since no residual term is present such as in the \textit{small} and \textit{psmall} estimator (equation \ref{eq:pointest_2p_small} and \label{eq:pest_2p_psmall}) and thus look similar to the synthetic estimators (equation \ref{eq:pointest_2p_reg_synth}, \ref{eq:pointest_2p_reg_synth}, \ref{eq:var_2p_reg_synth}, \ref{eq:var_2p_reg_psynth}). However, note that by a decomposition of $\hat{\pmb{\theta}}_{s_2}$ it can be shown that the residual correction term is indeed included in the regression coefficient $\hat{\pmb{\theta}}_{s_2}$ \citep{mandallaz2016} and thus the estimator is design-unbiased.
